The Array of Long-Baseline Antennas for Taking Radio Observations from the Sub-Antarctic

### 2020-06-15 albatros_paper.tex
HCC tweaked correlator plot and text

The autonomous station can be configured to record 1-, 2-, or 4-bit
baseband.  Signal fidelity improves with higher numbers of bits, but
the penalty is increased data volume.  Here we discuss the prospects
of employing 1-bit correlation for the final \albatros\ array to keep
the baseband data volume to a minimum.  While the basics of 1-bit
correlation are well known in the limit where the signal level is much
lower than the noise, \albatros\ may well be in the high-signal
regime.  The fundamental output of a 1-bit correlator for real data is
\begin{equation}
  x_{ij} \equiv \left < \tilde{E}_i \tilde{E}_j \right >.
\end{equation}
Here $\tilde{E}_i$ is the quantized version of the underlying electric
field $E_i$ for antenna index $i$:
\begin{equation}
  \tilde{E}_i=
  \begin{cases}
    1 & \text{if}\ E_i > 0 \\
    -1 & \text{if}\ E_i < 0.
  \end{cases}
\end{equation}
Complex data can be handled as the combination of real components.
The output is nonlinear in the underlying true signal and noise
levels; in the limit of perfectly correlated electric fields, $x_{ij}$
saturates at unity.  Recovering the true sky signals requires
inverting this nonlinear relation using the Van Vleck corrections.
For a 2-level correlator, the relationship between $x_{ij}$ and the
sky signals is~\citep{1989ASPC....6...59D, 1446497}
\begin{eqnarray}
\label{eqn:1bit_output}
\left < \sin\left(\frac{\pi}{2}x_{ij}\right)\right > = \frac{V_{ij}}{\sqrt{V_{ij}+N_i}\sqrt{V_{ij}+N_j}},
\end{eqnarray}
where $V_{ij}$ is the true sky visibility, and $V_{ij}+N_{i}$ and
$V_{ij}+N_{j}$ are the total, signal-plus-noise, power levels measured
by antennas $i$ and $j$, respectively.  Inverting this relation to
obtain $V_{ij}$ requires knowing the total power levels, which cannot
be determined from 1-bit data alone.  However, as described in
\S\ref{s:autonomous_backend}, the SNAP board calculates and records
auto-spectrum data products, which provide power level information on
few-second time scales.  Changes in the power levels occur on
significantly longer time scales, so the true sky visibilities can be
derived by combining cross-correlation data between different antenna
stations and the auto-spectrum data from each individual station.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.7\linewidth]{Figures/corr_efficiency/corr_efficiency.png}
    \caption{Correlator efficiency $\epsilon_{corr}$ for a 1-bit correlator as a function of signal-to-noise ratio (SNR).  To get the expected SNR from a 1-bit correlator, multiply the SNR derived from the radiometer equation by $\epsilon_{corr}$.  For low SNR, the efficiency is $2/\pi \sim 0.64$, and it drops monotonically as the SNR increases.  The drop is relatively gentle, only changing by a factor of $\sim 2$ for SNR=2.  For baseline separations of many wavelengths, the correlated component of the electric field is likely to be a small fraction of the total power, so ALBATROS is unlikely to lose more than a factor of $\sim 2$ by using 1-bit correlation.}
    \label{fig:1bit_efficiency}
    \end{center}
\end{figure}

One might worry that at high signal levels, the saturation of the 1-bit cross correlation would lead to the noise exploding.  In practice the rolloff in sensitivity is relatively mild, especially in the case of long baselines.  It can be expressed in terms of the correlator efficiency $\epsilon_{corr}$, which is defined to be the ratio of the measured signal-to-noise ratio to the ideal (infinite precision) SNR.  As seen in Figure \ref{fig:1bit_efficiency}, $\epsilon_{corr}$ only drops by a factor of $\sim$2 as the signal power goes from 0 to a few times the noise power.  We stress that the signal level in question is the part that correlates between the two antennas.  In the case of long baselines (where long means the fringe spacing is small compared to the primary beam/antenna response pattern) that are not dominated by a single bright source, the signal power will always be smaller than the noise power, and so $\epsilon_{corr}$ is unlikely to drop below $\sim$0.5\footnote{A quick estimate can be made by analogy to the behavior of dishes in the UV plane in the flat sky approximation.  While all power in the UV plane (plus any system noise) goes into an individual antenna's electric field, the correlated part only has contribution from the area in the UV plane within the width of the UV-space primary beam of the UV-space coordinate.  If the sky can be described as a Gaussian random field, in the most pessimistic case this is roughly the aperture filling factor of the antenna pair.  In our case, where the antennas are approximately dipolar, the filling factor will be the square of the wavelength over the baseline length.  As long as the baselines are several wavelengths or longer, the correlated power will be small, and the correlator efficiency will be close to the ideal 1-bit value of $\frac{2}{\pi}$.}.

>>>>

The autonomous station can be configured to record 1-, 2-, or 4-bit
baseband.  Signal fidelity improves with higher numbers of bits, but
the penalty is increased data volume.  Here we discuss the prospects
of employing 1-bit correlation for the final \albatros\ array to keep
the baseband data volume to a minimum.  While the basics of 1-bit
correlation are well known in the limit where the signal level is much
lower than the noise, \albatros\ may be in a regime where the signal
level is non-negligible.  The fundamental output of a 1-bit correlator
for real data is
\begin{equation}
  x_{ij} \equiv \left < \tilde{E}_i \tilde{E}_j \right >.
\end{equation}
Here $\tilde{E}_i$ is the quantized version of the underlying electric
field $E_i$ for antenna index $i$:
\begin{equation}
  \tilde{E}_i=
  \begin{cases}
    1 & \text{if}\ E_i > 0 \\
    -1 & \text{if}\ E_i < 0.
  \end{cases}
\end{equation}
Complex data can be handled as the combination of real components.
The output is nonlinear in the underlying true signal and noise
levels; in the limit of perfectly correlated electric fields, $x_{ij}$
saturates at unity.  Recovering the true sky signals requires
inverting this nonlinear relation using the Van Vleck
corrections~\citep{1446497}.  For a 2-level correlator, the
relationship between $x_{ij}$ and the sky signals
is~\citep{1989ASPC....6...59D}
\begin{eqnarray}
\label{eqn:1bit_output}
\left < \sin\left(\frac{\pi}{2}x_{ij}\right)\right > = \frac{V_{ij}}{\sqrt{V_{ij}+N_i}\sqrt{V_{ij}+N_j}},
\end{eqnarray}
where $V_{ij}$ is the true sky visibility, and $V_{ij}+N_{i}$ and
$V_{ij}+N_{j}$ are the total, signal-plus-noise, power levels measured
by antennas $i$ and $j$, respectively.  Inverting this relation to
obtain $V_{ij}$ requires knowing the total power levels, which cannot
be determined from 1-bit data alone.  However, as described in
\S\ref{s:autonomous_backend}, the SNAP board calculates and records
auto-spectrum data products, which provide power level information on
few-second time scales.  Changes in the power levels occur on
significantly longer time scales, so the true sky visibilities can be
derived by combining cross-correlation data between different antenna
stations and the auto-spectrum data from each individual station.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.7\linewidth]{Figures/corr_efficiency/corr_efficiency.png}
    \caption{Correlator efficiency $\epsilon$ for a 1-bit correlator
      as a function of signal-to-noise ratio (SNR).  The efficiency is
      roughly 0.64 at low SNR and drops monotonically as the SNR
      increases.  The decrease in $\epsilon$ is relatively gentle,
      changing by only a factor of 2 for SNR $\sim2$.}
    \label{fig:1bit_efficiency}
    \end{center}
\end{figure}

% To get the expected SNR from a 1-bit correlator, multiply the SNR
% derived from the radiometer equation by $\epsilon_{corr}$.

To assess the loss of sensitivity as the signal level increases, we
use \autoref{eqn:1bit_output} to derive the correlator efficiency
$\epsilon$, which is the ratio of measured signal-to-noise ratio (SNR)
to the ideal, infinite-precision SNR.  \autoref{fig:1bit_efficiency}
illustrates that the decrease in $\epsilon$ as a function of SNR is
relatively shallow.  When the SNR is near zero, $\epsilon \approx
2/\pi = 0.64$, and if the signal level increases to two times the
noise level, $\epsilon$ decreases by only a factor of $\sim2$.  The
final \albatros\ array will have baselines that are long relative to
the observing wavelength, and the correlated component of the electric
field (which sets the relevant signal level) is likely a small
fraction of the total received power.  This qualitative argument can
by made by considering antennas in the $uv$ plane in the flat-sky
approximation.  Any individual antenna receives all power present in
the $uv$ plane (plus system noise), but the correlated contribution
comes only from the portion of the $uv$ plane that is encompassed by
the primary beam at the antenna's $uv$-space coordinate.  Assuming
that the sky can be described as a Gaussian random field, the maximum
correlated signal between two antennas is determined by the aperture
filling factor of that pair.  For the case of dipole antennas, the
filling factor is the square of the wavelength divided by the baseline
length.  Therefore, in the long-baseline limit, the correlated signal
is small in comparison to the total power received, and the correlator
efficiency is unlikely to decrease significantly from the ideal
$2/\pi$ value.

% One might worry that at high signal levels, the saturation of the
% 1-bit cross correlation would lead to the noise exploding.  In
% practice the rolloff in sensitivity is relatively mild, especially in
% the case of long baselines.  It can be expressed in terms of the
% correlator efficiency $\epsilon_{corr}$, which is defined to be the
% ratio of the measured signal-to-noise ratio to the ideal (infinite
% precision) SNR.  As seen in Figure \ref{fig:1bit_efficiency},
% $\epsilon_{corr}$ only drops by a factor of $\sim$2 as the signal
% power goes from 0 to a few times the noise power.  We stress that the
% signal level in question is the part that correlates between the two
% antennas.  In the case of long baselines (where long means the fringe
% spacing is small compared to the primary beam/antenna response
% pattern) that are not dominated by a single bright source, the signal
% power will always be smaller than the noise power, and so
% $\epsilon_{corr}$ is unlikely to drop below $\sim$0.5\footnote{A quick
%   estimate can be made by analogy to the behavior of dishes in the UV
%   plane in the flat sky approximation.  While all power in the UV
%   plane (plus any system noise) goes into an individual antenna's
%   electric field, the correlated part only has contribution from the
%   area in the UV plane within the width of the UV-space primary beam
%   of the UV-space coordinate.  If the sky can be described as a
%   Gaussian random field, in the most pessimistic case this is roughly
%   the aperture filling factor of the antenna pair.  In our case, where
%   the antennas are approximately dipolar, the filling factor will be
%   the square of the wavelength over the baseline length.  As long as
%   the baselines are several wavelengths or longer, the correlated
%   power will be small, and the correlator efficiency will be close to
%   the ideal 1-bit value of $\frac{2}{\pi}$.}.
